---
title: "Predicting Exercise Activity Quality"
author: "Jim Quinn"
date: "September 26, 2015"
output: html_document
---

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

## Goal
Use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Outcome Classification
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).
Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate.

## Executive Summary
Provided training data was explored, cleaned and analyzed with the goal of creating a machine learning algorithm to be used for prediction of a testing data set. The selected prediction model was created using  the carot package Random Forest function. The resulting model produced 100% accuracy after submitting predictions.
In an effot to create this analysis as Reproducible Research, the analysis is as follows.

## Analysis 
#### The following provide code and comments describing how the analysis was performed.

### Load the Libraries
```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
```

### Get the Data
```{r, echo=TRUE}
fileURL1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileURL2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if (!file.exists("data")) {
  dir.create("data")
}
if (!file.exists("./data/pml-training.csv")) {
  download.file(fileURL1, destfile = "./data/pml-training.csv")
}
# Read training data with defined NA Strings
trainingData <- read.csv("./data/pml-training.csv", na.strings = c("NA","#DIV/0!",""))
if (!file.exists("./data/pml-testing.csv")) {
  download.file(fileURL2, destfile = "./data/pml-testing.csv")
}
# Read testing data with defined NA Strings
testingData <- read.csv("./data/pml-testing.csv", na.strings = c("NA","#DIV/0!",""))
```

### Clean Data
#### Choose columns to exclude from analysis
```{r}
# Exclude Columns with strings found in excludeColumns
excludeColumns <- grep("X|name|timestamp|window", colnames(trainingData), value = FALSE) 
cleanTrainingData <- trainingData[,-excludeColumns] # 160 columns now 153
# Remove columns that contain more than 95% NAs
naRate <- apply(cleanTrainingData, 2, function(x) sum(is.na(x)))/nrow(cleanTrainingData)
cleanTrainingData <- cleanTrainingData[!(naRate > 0.95)] # 153 columns reduced to 53
cleanColumnNames <- names(cleanTrainingData)
# Add character column "classe" to tesingData, simplifying subsetting of columns in sync with cleanTrainingData.
testingData$classe <- ""
cleanTestingData <- testingData[,cleanColumnNames]
```

### Partition Training Data 
#### Split provided training data into two sets - 75% myTrainingSet and 25% myTrainingTestSet
```{r}
set.seed(13234)
inTrain <- createDataPartition(y = cleanTrainingData$classe, p = 0.75, list = FALSE)
myTrainingSet <- cleanTrainingData[inTrain, ] 
myTrainingTestSet <- cleanTrainingData[-inTrain, ]
```

### Visulize Data - Rpart Decision Tree
```{r}
model1 <- rpart(classe ~ ., data = myTrainingSet, method = "class")
prp(model1)
```

### Principal Components Analysis (PCA)
#### Using preProcess() determined that 25 components are required to achieve 95% confidence threshold
#### Using 25 for pcaComp, determine which 25 variables should be included in analysis
```{r}
thresh95 <- preProcess(myTrainingSet[,1:52],method = "pca",thresh = .95)
print(thresh95$numComp - 1) # 25 Components required for 95% threshold
pca <- preProcess(myTrainingSet[,1:52],method = "pca",pcaComp = 25) 
pcaTraining <- predict(pca,myTrainingSet[,1:52])
```
 
### Create Model using Random Forests
```{r}
modelFit <- randomForest(myTrainingSet$classe ~ .,   data = pcaTraining, do.trace = FALSE)
print(modelFit)
```

### Cross-Validation
#### In random forests, there is no need for cross-validation or a separate test set to get an unbiased estimate of the test set error. It is estimated internally, during the run.
#### OOB (out-of-bag) estimate is 2.46%

### Prediction Using Data Partitioned for Testing Split from Provided Training Data
```{r}
testingPredict <- predict(pca,myTrainingTestSet[,1:52])
confusionMatrix(myTrainingTestSet$classe,predict(modelFit,testingPredict))
```
#### Accuracy is 97.7% - which means out of sample error is 2.3%.


## Applying machine learning algorithm to the 20 test cases
```{r}
testDataPrediction <- predict(pca,cleanTestingData[,1:52])
cleanTestingData$classe <- predict(modelFit,testDataPrediction)
answers <- predict(modelFit,testDataPrediction)
```
### Submitted all test case predictions, recieving confirmation of 100% accuracy!

## Refernces/Citations

Data set for this analysis provided by:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 
